Quantifying Limits to Detection of Early Warning for Critical Transitions

### 2012-02-14 boettiger.tex
Merged Alan's edits

However, for many ecological systems specific models are not available and general approaches are needed~\citep{Scheffer2009}
that do not depend on estimating the parameters of a known model of a specific system. 
This has lead to a variety of approaches based on summary statistics
~\citep[\emph{e.g.}][]{Carpenter2006, Held2004, Dakos2008, Guttal2008, Biggs2009, Carpenter2011, Seekell2011}
that look for generic signs of impending regime shifts.  
Our goal here is to extend earlier work by providing much more precise estimates 
of the ability of different statistics to signal impending regime shifts 
and to develop new approaches that both are more efficient 
and also lay bare some of the important assumptions underlying attempts 
to find general warning signs of regime shifts.  
We caution against paradigms that are not useful for capturing uncertainty in
a model-selection based approach, such as information criteria.

We begin our approach by first discussing the limitations of current approaches that rely on summary statistics
and provide a clear description of assumptions through the introduction of a model based approach to detect early warning signals.
We then illustrate how stochastic differential equation (SDE) models can be used
to reflect the uncertainty inherent in the detection of early warning signals.
Finally we suggest receiver-operating characteristics~\citep{Green1989, Keller2009}
as a powerful way to illustrate the sensitivity of different data sets
and different indicators have in detecting early warning signals and use this to explore a number of examples.

\section{The summary statistics approach}
Foundational work on early warning signals has operated under the often-implicit assumption
that the system dynamics contain a saddle-node bifurcation by
looking for patterns that are associated with this kind of transition.
Typical patterns include an increasing trend in a summary statistic such as
variance~\citep{Carpenter2006}, autocorrelation~\citep{Held2004, Dakos2008},
skew~\citep{Guttal2008}, spectral ratio~\citep{Biggs2009}.
While attractive for their simplicity, such approaches are subject to numerous challenges.
In this paper we argue for a model-based approach,
and describe how this can be done in a way that best addresses these difficulties.
We begin by enumerating several of the difficulties encountered in approaches lacking an explicit model.

@@

\subsubsection*{Hidden assumptions}
The underlying assumption that the system contains a saddle-node bifurcation,
can be easily overlooked in common summary-statistics based approaches.
For instance, variance may increase for reasons that
do not signal an approaching transition~\citep{Schreiber2003, Schreiber2008}.
Alternatively, variance may not increase as a bifurcation is approached~\citep{Livina2012, Dakos2011a}.
Some classes of sudden transitions may exhibit no warning signals~\cite{Hastings2010}.
Like saddle-node bifurcations, transcritical bifurcations involve an eigenvalue passing through zero,
and exhibit the patterns of critical slowing down and increased variance~\citep{Drake2010}.  
However, transcritical bifurcations are not ``blue-sky'' transitions that have made critical transitions so worrisome --
they are characterized by a steady approach towards the alternative stable state, 
such as a population size gradually approaching zero.  
While no approach will be applicable to all classes of sudden transitions,
it is certainly still useful to have an approach that detect transitions driven by
saddle node bifurcations, which have been found in many contexts~\citep[\emph{e.g.}, see][]{Scheffer2001}.

%Differing causes of the transition.
Even when we can exclude or ignore such dynamics and
restrict ourselves to systems that can support a saddle-node bifurcation,
these approaches must further assume that a changing parameter
has brought the system closer to the bifurcation.
This excludes at least three alternative explanations for the transition.
The first is that a large perturbation of the system state
has moved the system into the alternative basin of attraction~\citep{Scheffer2001, Scheffer2001}.
This is an exogenous forcing that does not arise from the system dynamics, so it is not the kind of event we can expect to forecast.
(An example might be a sudden dramatic increase in fishing effort that pushes the population past a threshold.)
The second scenario is a purely noise-induced transition, a chance fluctuation that happens to carry the system across the boundary.~\citep{Ditlevsen2010}
As discussed in \citet{Livina2012}, such noise induced transitions cannot be predicted through early warning signals.
The third scenario is that the system does pass through a saddle-node bifurcation,
but rather than gradually and monotonically approaching the critical point, the
bifurcation parameter moves in a rapid or highly non-linear way, making the detection of any gradual trend impossible.

@@

\subsubsection*{Arbitrary windows}
In addition to the assumption of a saddle node bifurcation, 
the calculation of these statistics is subject to several additional arbitrary choices.
A basic difficulty arises from the need to assume a time-series is ergodic: 
that averaging over time is equivalent to averaging over replicate realizations,
while trying to test if it is not.
Theoretically, the increasing trend in variance, autocorrelation, or other statistics
is something that would be measured across an ensemble -- across replicates. %% citation?
As true replicates are seldom available in systems for which developing warning signals would be most desirable,
these methods average across a single replicate using a moving window in time.
The selection of the size of this window and whether and by how much to overlap consecutive windows
varies across the literature.
\citet{Lenton2012} demonstrates that these differences can influence the results,
and that the different choices each carry advantages and disadvantages.

@@

Further, certain statistics such as the critical slowing down measured in autocorrelation
require data that is evenly sampled in time.
Interpolating from existing data to create evenly spaced points is particularly problematic,
as this introduces an artificial autocorrelation into the data.

@@

\section{A model based approach}
Model-based approaches are beginning to play a larger role in early warning signals,
though we have not as yet seen the direct fitting and simulation of models to compare hypotheses.
While choosing appropriate models without system-specific knowledge is challenging,
much can be accomplished by framing the implicit assumptions into equations.
\citet{Lade2011} introduces the idea of generalized models for early warning signals, and
\citet{Kuehn2011} presents normal forms for bifurcation processes that can give rise to critical transitions.
\citet{Carpenter2011e} and \citet{Dakos2011a} start by assuming the dynamics obey a generic
stochastic differential equation (SDE), but use this only to derive or define the summary statistics of interest.

@@

\subsection*{Models}
The normal form~\citep{Guckenheimer1983, Kuehn2011} for the saddle-node bifurcation is
\begin{equation}
\frac{\ud x}{\ud t} = r_t- x^2.
\label{saddle-node}
\end{equation}
where $x$ is the state variable and $r_t$ our bifurcation parameter,
which may be slowly varying in time and
consequently moving the system closer to a critical transition or regime shift~\citep{Scheffer2009}.
Transforming this canonical form to allow for an arbitrary mean $\theta$,
the bifurcation looks like \( dx/dt = r_t- (\theta-x)^2 \), with fixed point \(\hat x = \sqrt{r_t} +\theta =: \phi(r_t)\).
We expand around the fixed point and express as a stochastic differential equation~\citep[\emph{e.g.}][]{Gardiner2009}:

@@

where $B_t$ is the standard Brownian motion. 
This expression captures the behavior of the system near the stable point as it approaches 
%It is also possible to derive this same expression from the master equation
%of various stochastic birth/death (or gains \& losses) models
%that contain saddle-node bifurcations, as we show in the appendix.
%Alternate dynamical forms are possible but will not differ substantially.

As we discuss above, in this paradigm we must include an assumption on how the bifurcation parameter, $r_t$, is changing.
We assume a gradual, monotonic change which we approximate to first order:
\begin{equation}
r_t = r_0 - m t.
\label{R_t}
\end{equation}
Detecting accelerating or otherwise nonlinear approaches to the bifurcation will generally require more power.
When the system is stable, $r_t$ is constant and Equation~\eqref{LSN} will reduce to a simple Ornstein-Uhlenbeck process,
\begin{equation}
\ud X_t = r (\theta - X_t) \ud t + \sigma \ud B_t \label{OU}
\end{equation}
This is the continuous time analog of the first-order autoregressive model considered as a null model elsewhere~\citep[\emph{e.g.}][]{Dakos2008, Guttal2008a}.

@@

\subsection*{Comparing Models}
Likelihood methods form the basis of much of modern statistics
in both Frequentist and Bayesian paradigms.
The ability evaluate likelihoods directly by computation has made it
possible to treat cases that do not conform to traditional assumptions more directly.
The basis of likelihood comparisons has its roots in the Neyman-Pearson Lemma,
which essentially asserts that comparing likelihoods is the most powerful test
of a choice between two hypotheses~\citep{Neyman1933}, and motivates
tests from the simple likelihood ratio test up through modern model adequacy methods.

@@

The ROC curves for each of these data sets is plotted in Figure~\ref{fig:rocdata}.  
While differences in the rate at which the system approaches a transition will also 
improve the ratio of true positives to false positives, 
here we see the best-sampled data set, Glaciation, with 121 points, 
also has the clearest signal with no observed errors in the 500 replicates of each type.   
Comparing the chemostat and simulation curves illustrate how the trade-off between
false positives and true positives can vary between data.  The chemostat signal, 
which estimates a relatively rapid rate of change but has less data, captures a higher
rate of true positives for a given rate of false positives than the simulation data
set with a weaker rate of change but more data, for false positive rates above 20\%.  
However, the simulated set with more data performs better if lower false-positive rates
are desired.

@@

\section{Discussion}
By assuming the underlying model corresponds to a saddle-node bifurcation,
our analysis presents a ``best-case scenario'' for both the summary statistics and the likelihood-based approach.
Other literature has already begun to address the additional challenges posed when the underlying
dynamics do not correspond to these models~\citep{Hastings2010}.
Our results illustrate that even in this best-case scenario,
reliable identification of warning signals from summary statistics can be difficult.
The extent of this difficulty becomes clear from the ROC curves of these statistics,
where a 5\% false positive rate often corresponds to only a 5\% true positive rate,
performing no better than the flip of a coin. %%%% EEEK are you really going to say that?!?!
By estimating the ROC curve for a given set of data,
we can better avoid applying warning signals in cases of inadequate power.
By taking advantage of the assumptions being made to write down a specific likelihood function,
we can develop approaches that get the most information from the data available.

>>>>

However, for many ecological systems specific models are not available and general approaches are needed~\citep{Scheffer2009}
that do not depend on estimating the parameters of a known model of a specific system. 
This has lead to a variety of approaches based on summary statistics
~\citep[\emph{e.g.}][]{Carpenter2006, Held2004, Dakos2008, Guttal2008, Biggs2009, Carpenter2011, Seekell2011}
that look for generic signs of impending regime shifts.  
Our goal here is to extend earlier work by providing much more precise estimates 
of the ability of different statistics to signal impending regime shifts 
and to develop new approaches that both are more efficient 
and also lay bare some of the important assumptions underlying attempts 
to find general warning signs of regime shifts.

We begin our approach by first discussing the limitations of current approaches that rely on summary statistics
and provide a clear description of assumptions through the introduction of a model based approach to detect early warning signals.
We then illustrate how stochastic differential equation (SDE) models can be used
to reflect the uncertainty inherent in the detection of early warning signals. We caution against paradigms that are not useful for capturing uncertainty in
a model-selection based approach, such as information criteria.
Finally we use receiver-operating characteristics~\citep{Green1989, Keller2009}
as a powerful way to illustrate the sensitivity of different data sets
and different indicators have in detecting early warning signals and use this to explore a number of examples.

\section{The summary statistics approach}
Foundational work on early warning signals has operated under the often-implicit assumption
that the system dynamics contain a saddle-node bifurcation by
looking for patterns that are associated with this kind of transition.
Typical patterns include an increasing trend in a summary statistic such as
variance~\citep{Carpenter2006}, autocorrelation~\citep{Held2004, Dakos2008},
skew~\citep{Guttal2008}, spectral ratio~\citep{Biggs2009}.
While attractive for their simplicity, such approaches must confront numerous challenges.
In this paper we argue for a model-based approach,
and describe how this can be done in a way that best addresses these difficulties.
We begin by enumerating several of the difficulties encountered in approaches lacking an explicit model.

@@

\subsubsection*{Hidden assumptions}
The underlying assumption that the system contains a saddle-node bifurcation
can be easily overlooked in common summary-statistics based approaches.
For instance, variance may increase for reasons that
do not signal an approaching transition~\citep{Schreiber2003, Schreiber2008}.
Alternatively, variance may not increase as a bifurcation is approached~\citep{Livina2012, Dakos2011a}.
Some classes of sudden transitions may exhibit no warning signals~\cite{Hastings2010}.
Like saddle-node bifurcations, transcritical bifurcations involve an eigenvalue passing through zero,
and exhibit the patterns of critical slowing down and increased variance~\citep{Drake2010}.  
However, transcritical bifurcations are not ``blue-sky'' transitions that have made critical transitions so worrisome --
they are characterized by a steady approach towards the alternative stable state, 
such as a population size gradually approaching zero.  
While no approach will be applicable to all classes of sudden transitions,
it is certainly still useful to have an approach that detect transitions driven by
saddle node bifurcations, which have been found in many contexts~\citep[\emph{e.g.}, see][]{Scheffer2001}.

%Differing causes of the transition.
Even when we can exclude or ignore other dynamics and
restrict ourselves to systems that can produce a saddle-node bifurcation,
 approaches based on critical slowing down or rising variance (REFS) must further assume that a changing parameter
has brought the system closer to the bifurcation.
This assumption excludes at least three alternative explanations for the transition.
The first possibility is that a large perturbation of the system state
has moved the system into the alternative basin of attraction~\citep{Scheffer2001, Scheffer2001}.
This is an exogenous forcing that does not arise from the system dynamics, so it is not the kind of event we can expect to forecast.
(An example might be a sudden dramatic increase in fishing effort that pushes the population past a threshold.)
The second scenario is a purely noise-induced transition, a chance fluctuation that happens to carry the system across the boundary~\citep{Ditlevsen2010}.
As discussed in \citet{Livina2012}, such noise induced transitions cannot be predicted through early warning signals.
The third scenario is that the system does pass through a saddle-node bifurcation,
but rather than gradually and monotonically approaching the critical point, the
bifurcation parameter moves in a rapid or highly non-linear way, making the detection of any gradual trend impossible.

@@

\subsubsection*{Arbitrary windows}
In addition to the assumption of a saddle node bifurcation, 
the calculation of statistics that would be used to detect an impending transition is subject to several additional arbitrary choices.
A basic difficulty arises from the need to assume a time-series is ergodic: 
that averaging over time is equivalent to averaging over replicate realizations,
while trying to test if it is not.
Theoretically, the increasing trend in variance, autocorrelation, or other statistics
is something that would be measured across an ensemble -- across replicates. %% citation?
As true replicates are seldom available in systems for which developing warning signals would be most desirable,
these methods average across a single replicate using a moving window in time.
The selection of the size of this window and whether and by how much to overlap consecutive windows
varies across the literature.
\citet{Lenton2012} demonstrates that these differences can influence the results,
and that the different choices each carry advantages and disadvantages.

@@

Further, certain statistics such as the critical slowing down measured by autocorrelation
require data that is evenly sampled in time.
Interpolating from existing data to create evenly spaced points is particularly problematic,
as this introduces an artificial autocorrelation into the data.

@@

\section{A model based approach}
Model-based approaches are beginning to play a larger role in early warning signal detection,
though we have not as yet seen the direct fitting and simulation of models to compare hypotheses.
While choosing appropriate models without system-specific knowledge is challenging,
much can be accomplished by framing the implicit assumptions into equations.
\citet{Lade2011} introduces the idea of generalized models for early warning signals, and
\citet{Kuehn2011} presents normal forms for bifurcation processes that can give rise to critical transitions.
\citet{Carpenter2011e} and \citet{Dakos2011a} start by assuming the dynamics obey a generic
stochastic differential equation (SDE), but use this only to derive or define the summary statistics of interest.

@@

\subsection*{Models}
The normal form~\citep{Guckenheimer1983, Kuehn2011} for the saddle-node bifurcation is
\begin{equation}
\frac{\ud x}{\ud t} = r_t- x^2.
\label{saddle-node}
\end{equation}
where $x$ is the state variable and $r_t$ our bifurcation parameter,
which may be slowly varying in time and
consequently moving the system closer to a critical transition or regime shift~\citep{Scheffer2009}.
Transforming this canonical form to allow for an arbitrary mean in the state variable $\theta$,
the bifurcation looks like \( dx/dt = r_t- (\theta-x)^2 \), with fixed point \(\hat x = \sqrt{r_t} +\theta =: \phi(r_t)\).
We expand around the fixed point and express as a stochastic differential equation~\citep[\emph{e.g.}][]{Gardiner2009}:

@@

where $B_t$ is the standard Brownian motion. 
This expression captures the behavior of the system near the stable point as it approaches the bifurcation.
%It is also possible to derive this same expression from the master equation
%of various stochastic birth/death (or gains \& losses) models
%that contain saddle-node bifurcations, as we show in the appendix.
%Alternate dynamical forms are possible but will not differ substantially.

As we discuss above, in this paradigm we must include an assumption on how the bifurcation parameter, $r_t$, is changing.
We assume a gradual, monotonic change which we approximate to first order:
\begin{equation}
r_t = r_0 - m t.
\label{R_t}
\end{equation}
Detecting accelerating or otherwise nonlinear approaches to the bifurcation will generally require more power.
When the system is stable, $r_t$ is constant ($m=0$) and Equation~\eqref{LSN} will reduce to a simple Ornstein-Uhlenbeck process,
\begin{equation}
\ud X_t = r (\theta - X_t) \ud t + \sigma \ud B_t \label{OU}
\end{equation}
This is the continuous time analog of the first-order autoregressive model considered as a null model elsewhere~\citep[\emph{e.g.}][]{Dakos2008, Guttal2008a}.

@@

\subsection*{Comparing Models}
Likelihood methods form the basis of much of modern statistics
in both Frequentist and Bayesian paradigms.
The ability to evaluate likelihoods directly by computation has made it
possible to treat cases that do not conform to traditional assumptions more directly.
The basis of likelihood comparisons has its roots in the Neyman-Pearson Lemma,
which essentially asserts that comparing likelihoods is the most powerful test
of a choice between two hypotheses~\citep{Neyman1933}, and motivates
tests from the simple likelihood ratio test up through modern model adequacy methods.

@@

The ROC curves for each of these data sets are plotted in Figure~\ref{fig:rocdata}.  
While differences in the rate at which the system approaches a transition will also 
improve the ratio of true positives to false positives, 
here we see the best-sampled data set, Glaciation, with 121 points, 
also has the clearest signal with no observed errors in the 500 replicates of each type.   
Comparing the chemostat and simulation curves illustrate how the trade-off between
false positives and true positives can vary between data.  The chemostat signal, 
which estimates a relatively rapid rate of change but has less data, captures a higher
rate of true positives for a given rate of false positives than the simulation data
set with a weaker rate of change but more data, for false positive rates above 20\%.  
However, the simulated set with more data performs better if lower false-positive rates
are desired.

@@

\section{Discussion}

The challenge of determining early warning signs for impending possible regime shifts requires real attention to the underlying statistical issues and other assumptions.  Doing this, does, however, open up new possibilities for asking what the goal of detection should be, and for clearly identifying underlying assumptions. We consider alternative approaches based either on summary statistics or a likelihood based model choice.  By assuming the underlying model corresponds to a saddle-node bifurcation,
our analysis presents a ``best-case scenario'' for both  summary statistic and  likelihood-based approaches.
Other literature has already begun to address the additional challenges posed when the underlying
dynamics do not correspond to these models~\citep{Hastings2010}.
Our results illustrate that even in this best-case scenario,
reliable identification of warning signals from summary statistics can be difficult.

Asa way to clearly illustrate the choices involved in looking for warning signals while avoiding false alarms, we introduce an approach based on receiver operator curves. These curves XXX. The extent of the difficulty in finding reliable indicators of impending regime shifts based on summary statistics. becomes clear from the ROC curves of these statistics,
where a 5\% false positive rate often corresponds to only a 5\% true positive rate,
performing no better than the flip of a coin. %%%% EEEK are you really going to say that?!?!
By estimating the ROC curve for a given set of data,
we can better avoid applying warning signals in cases of inadequate power.
By taking advantage of the assumptions being made to write down a specific likelihood function,
we can develop approaches that get the most information from the data available.

DISUCSS EXAMPLES HERE
